---
title: "albums_exploration"
format: html
editor: visual
---

```{r}
library(readr)
library(ggplot2)
library(dplyr)
```

```{r}
albums <- read_csv("data/albums.csv")
albums <- filter(albums, artist == c("kanye west", "kid cudi"))
```

# Difference between `geom_col` and `geom_bar`

## `geom_col()`

geom_col() simply says: "Use the values I provide as the bar heights." So a y-axis **must** be provided. If there are multiple bars per category, the bars become stacked by default.

```{r}
albums |>
  ggplot(aes(x=artist, y=rating, fill=project_name)) +
  geom_col()
```

If you don't want to stack the bars but have them side-by-side, you can use `position="dodge"`

```{r}
albums |>
  ggplot(aes(x=artist, y=rating, fill=project_name)) +
  geom_col(position="dodge")
```

## `geom_bar()`

For geom_bar(), the default behavior is to count the rows for each x value. It doesn't expect a y-value, since it's going to count that up itself. In fact, it will flag a warning if you give it one.

```{r}
albums |>
  ggplot(aes(x=artist, fill=project_name)) +
  geom_bar()
```

You can dodge the position to not stack but have bars side-by-side.

```{r}
albums |>
  ggplot(aes(x=artist, fill=project_name)) +
  geom_bar(position="dodge")
```

You can use the weight aesthetic to **sum** the weights you choose.

```{r}
albums |>
  ggplot(aes(x=artist, weight=rating)) +
  geom_bar()
```

## stat="identity"

Every layer has a `stat` "part" (statistical transformations). The stat part performs a computation on the data before it is displayed.

Note: **Stats determine *what* is displayed, not *how* it is displayed.** For example, if you add [`stat_density()`](https://ggplot2.tidyverse.org/reference/geom_density.html) to a plot, a kernel density estimation is performed, which *can be displayed* with the 'geom' part of a layer.

For many `geom_*()` functions, [`stat_identity()`](https://ggplot2.tidyverse.org/reference/stat_identity.html) is used, which performs no extra computation on the data. For a geom_bar the default stat is `count`, for geom_col it is `identity`.

You can define the stat part of a layer in different ways: for example, you can provide it to a geom as a string.

```{r}
albums |>
  ggplot(aes(x=artist, y=rating, fill=project_name)) +
  geom_bar(stat="identity", position="dodge")
```
