---
title: "albums_exploration"
format: html
editor: visual
---

```{r}
library(readr)
library(ggplot2)
library(dplyr)
```

```{r}
albums <- read_csv("data/albums.csv")
albums <- filter(albums, artist == "kanye west")
```

# Difference between geom_col and geom_bar

## `geom_col()`

geom_col() simply says: "Use the values I provide as the bar heights." So a y-axis **must** be provided. If there are multiple bars per category, the bars become stacked.

```{r}
albums |>
  ggplot(aes(x=artist, y=rating, fill=project_name)) +
  geom_col()
```

If you don't want to stack the bars but have them side-by-side, you can use `position="dodge"`

```{r}
albums |>
  ggplot(aes(x=artist, y=rating, fill=project_name)) +
  geom_col(position="dodge")
```

## `geom_bar()`

For geom_bar(), the default behavior is to count the rows for each x value. It doesn't expect a y-value, since it's going to count that up itself. In fact, it will flag a warning if you give it one, since it thinks you're confused. 

```{r}
albums |>
  ggplot(aes(x=artist, fill=project_name)) +
  geom_bar()
```

You can dodge the position to not stack but have bars side-by-side.

```{r}
albums |>
  ggplot(aes(x=artist, fill=project_name)) +
  geom_bar(position="dodge")
```

You can use the weight aesthetic to sum the weights you choose.

```{r}
albums |>
  ggplot(aes(x=artist, weight=rating)) +
  geom_bar()
```


How aggregation is to be performed is specified as an argument to geom_bar(), which is stat = "count" for the **default** value.

If you explicitly say stat = "identity" in geom_bar(), you're telling ggplot2 to skip the aggregation and that you'll provide the y values. This mirrors the natural behavior of geom_col().

## stat="identity"

```{r}
albums |>
  ggplot(aes(x=artist, y=rating, fill=project_name)) +
  geom_bar(stat="identity", position="dodge")
```
