---
title: "Bar chart"
format: html
editor: visual
---

```{r}
#| include: false
library(readr)
library(ggplot2)
library(dplyr)
```

```{r}
#| include: false
albums <- read_csv("data/albums.csv")
top_artists <- albums |>
  count(artist, sort = TRUE) |>
  slice_head(n = 3) |>
  pull(artist)
albums <- filter(albums, artist %in% top_artists)
```

## Sample data

```{r}
#| echo: false
knitr::kable(albums)
```

### 1. x:disc y:none agg:count

Uses count(x) by default for y axis

```{r}
albums |>
  ggplot(aes(x=artist)) +
  geom_bar()
```

### 2. x:none y:disc agg:count

Uses count(y) by default for x axis

```{r}
albums |>
  ggplot(aes(y=artist)) +
  geom_bar()
```

### 3. x:disc y:cont agg:mean

Aggregates y with mean per x

```{r}
albums |>
  ggplot(aes(x=artist, y=rating)) +
  geom_bar(stat="summary", fun="mean")
```

### 4. x:cont y:disc agg:mean

Aggregates x with mean per y (horizontal)

```{r}
albums |>
  ggplot(aes(x=rating, y=artist)) +
  geom_bar(stat="summary", fun="mean", orientation="y")
```

### 5. x:disc y:cont agg:none

a\. One bar per row, in dataset order, no grouping or reordering (Currently in adept)

```{r}
albums |>
  mutate(row = row_number()) |>
  ggplot(aes(x=factor(row), y=rating)) +
  geom_col(width=1, color="black") +
  scale_x_discrete(labels = albums$artist)
```

b\. By default it shows the sums of the y by stacking them up

```{r}
albums |>
  ggplot(aes(x=artist, y=rating)) +
  geom_col(color="black")
```

c\. With dodging (Probably a better default in adept?)\

```{r}
albums |>
  ggplot(aes(x=artist, y=rating, group=project_name)) +
  geom_col(position="dodge", color="black")
```

### 6. x:cont y:disc agg:none

```{r}
albums |>
  ggplot(aes(x=rating, y=artist, group=project_name)) +
  geom_col(position="dodge", color="black")
```
