---
title: "Box Plots"
format: html
editor: visual
---

::: {.callout.danger}
https://www.elastic.co/docs/reference/aggregations/search-aggregations-metrics-boxplot-aggregation#search-aggregations-metrics-boxplot-aggregation-approximation
:::

## Box Plots

```{r}
#| echo: false
library(ggplot2)
library(dplyr)
library(patchwork)
library(tidyr)

economics <- economics |>
  mutate(category = rep(c("A", "B", "C"), length.out = n()))
head(economics)
```

A boxplot fundamentally does two things:

- summarizes one **continuous variable** (five-number summary) \[measured axis\]

- positions boxes along the other**, categorical axis** (one box per level). \[positioning/grouping axis\]

In ggplot you can create a boxplot in multiple different ways:

- provide only a continuous (measured) axis: since ggplot detects no categorical axis, you get a single group

```{r}
continuous_single_aes_horizontal <- ggplot(economics, aes(pop)) + 
  geom_boxplot(outliers=TRUE)
continuous_single_aes_vertical <- ggplot(economics, aes(y=pop)) + 
  geom_boxplot(outliers=TRUE)

continuous_single_aes_horizontal + continuous_single_aes_vertical
```

- provide a continuous and a grouping axis: here `category` is discrete, so it defines the groups (one box per level), and `pop` (continuous) is what's summarized.

```{r}
classic_horizontal <- ggplot(economics, aes(x=category, y=pop)) + 
  geom_boxplot(outliers=TRUE)
classic_vertical <- ggplot(economics, aes(x=pop, y=category)) + 
  geom_boxplot(outliers=TRUE)

classic_horizontal + classic_vertical
```

Boxplots do not **need** a categorical variable at all. So, we should enable passing a single variable too (e.g just an x-axis or just a y-axis). Our code should figure out the type and what to display:

- 1 category, 0 continuous: no plot

- 1 category, 1 continuous: plot

- 0 category, 1 continuous: plot

### Box plot with a date

```{r}
ggplot(economics, aes(x=date)) + 
  geom_boxplot(outliers=TRUE)
```

```{r}
ggplot(economics, aes(x=date, y=category)) + 
  geom_boxplot(outliers=TRUE)
```

## Compatibility table

**Note**: axis can be flipped.

| X-axis     | Y-axis     | Supported (in ggplot) | Plot                          |
|------------------|------------------|------------------|------------------|
| category   | NA         | No                    | **not supported**             |
| category   | continuous | Yes                   | show boxplot per category     |
| continuous | NA         | Yes                   | show single boxplot           |
| date       | NA         | Yes\*                 | show single boxplot           |
| date       | category   | Yes\*                 | show one boxplot per category |
| date       | continuous | No                    | **not supported**             |

\*remember that dates are internally represented as numeric (i.e continuous)

## Open questions for adept

- do we want to support dates as continuous variable?

  - we could probably do it by simply adjusting the "type" of the axis to "time" in echarts, so that the labels are nicely rendered (TBD)

```{r}
p <- ggplot(mpg, aes(class, hwy))
p + geom_boxplot()
```

- how to handle long labels on the x: currently, we let echarts hide them on overlap, which is not great

  - but we will always hit a limit...

- do we want to show outliers? our current endpoint for the boxplot does not return outliers, so we would need to compute them ourself...

  - which would be very sub-optimal, because our endpoint only returns us the computed 5-summary statistic

  - to compute outliers, we would need to send filter queries again to search for data points in a column above and below a certain threshold...

```{r}
p <- ggplot(economics, aes(y=unemploy)) + 
  geom_boxplot()

layer_data(p, 1)

d <- layer_data(p, 1)
lab <- tidyr::pivot_longer(d[, c("x","ymin","lower","middle","upper","ymax")],
                           -x, values_to = "v")

p + geom_text(data = lab, aes(x = x, y = v, label = round(v, 1)),
              inherit.aes = FALSE, hjust = -0.3, size = 3)
```

## Five point summary calculation methods

Confirmed ggplot uses R-7 (linear interpolation, `type = 7`).

### Example

```         
pos:  1   2   3   4   5   6   7   8   9
val: 29  29  32  32  32  32  33  36  36
```

n = 9. The median (Q2) is the 5th value = **32**. Every tool agrees on it, because the median sits on a single unambiguous position for odd n.

### The problem

Q1 and Q3 are the 25th and 75th percentiles. These are *positions* along the sorted data ("a quarter of the way through 9 values"), not data points. That position generally falls *between* two observations, so each tool needs a rule for where exactly to place it and how to interpolate. There are about nine standard rules (R types 1–9) plus non-standard ones. Same data with a different rule leads to different Q1/Q3. The data doesn't change, only the placement convention does.

### What ggplot uses

`geom_boxplot` / `stat_boxplot` calls R's `stats::quantile` with **type 7** (the default; it's now exposable as a parameter, but 7 is the default). Type 7 is linear interpolation on the position `(n − 1)·p + 1`:

- Q1: (9−1)·0.25 + 1 = position 3 → **32**
- Q3: (9−1)·0.75 + 1 = position 7 → **33**

Here both land *exactly* on data points, so no interpolation is even needed, and ggplot returns **32 / 32 / 33**. Note this is type 7, deliberately *not* the Tukey "hinges" that base R's `boxplot()` uses. ggplot's docs call out that they differ slightly on small samples.

### What Elasticsearch uses

The percentiles aggregation does **not** use a textbook quantile type at all. By default it uses **TDigest**, an approximate algorithm built to scale to billions of values: it compresses the data into weighted "centroids" and interpolates percentiles between centroid centers. The practical consequences:

- For a tiny input like this one it's effectively exact, but the interpolation convention is its own (centroid-based), so its Q1/Q3 must not equal ggplot's type-7 values.

- It is explicitly **non-deterministic**, so the same data can yield slightly different numbers across runs, and accuracy is tunable via a `compression` parameter.

- An alternative, **HDR histogram**, can be requested for fixed-precision behavior instead.

### Conclusion

- ggplot = deterministic textbook interpolation (R type 7)
- Elasticsearch = an approximate, tunable, centroid-based estimator (TDigest) designed for scale.

**One flag worth checking.** The clean **30.5 / 34.5** is exactly the *exclusive / Tukey* textbook method (median of each half with the median excluded, R-6-style). That's a specific stats convention, and it is *not* what Elasticsearch's default TDigest computes. TDigest produces algorithm-specific values, not tidy textbook ones. If some tool gives 30.5 / 34.5, that tool is using the exclusive method.
