---
title: "histogram"
format: html
editor: visual
---

```{r}
#| echo: FALSE
#| warning: FALSE
library(ggplot2)
library(patchwork)
library(dplyr)
```

::: {.callout-note}
# Continuous variables
A variable is **continuous** if it can take any of an infinite set of ordered values. Numbers and date-times are two examples of continuous variables.
:::

```{r}
head(diamonds) |>
  select(carat, cut, price)
```


## Histograms

Histograms are used to visualize **continuous** distributions: a histogram divides the x-axis into *equally spaced* bins and then uses the height of a bar to display the number of observations that fall in each bin.

```{r}
  ggplot(diamonds, aes(x = carat)) +
    geom_histogram() +
    labs(title="Distribution of diamonds by quality")
```

### Bin width
Different bin widths can reveal different patterns, so you should always try different ones to understand the underlying shape of your data [^bins].

```{r}
#| out-width: "100%"
p <- function(bw) {
  ggplot(diamonds, aes(x = carat)) +
    geom_histogram(binwidth = bw) +
    labs(title = paste("binwidth =", bw))
}

p(0.1) + p(0.5) + p(1)
```

There are many different "rules" about how to choose bin width, each with pros and cons. You can read about them on [this Wikipedia page](https://en.wikipedia.org/wiki/Histogram#Number_of_bins_and_width).

A very commonly used histogram role is Freedman-Diaconis $(h = 2 \times \text{IQR} \times n^{-1/3})$, which uses the IQR and is therefore more robust to outliers.

```{r}
fd_bw <- function(x) { (max(x) - min(x)) / nclass.FD(x) }
ggplot(airquality) +
    aes(x = Temp) +
    geom_histogram(binwidth = fd_bw, color = "black", fill = "gray")
```

### Overlaying histograms
Sometimes you may want to display multiple distributions in the same plot.

```{r}
overlaying_histo <- ggplot(diamonds, aes(x=carat, fill=cut)) + 
  geom_histogram(show.legend = FALSE)
overlaying_histo
```

While this works, it can be difficult to interpret the result. Frequency polygons are often the better tool for this use case. `geom_freqpoly()` performs the same calculation as [`geom_histogram()`](https://ggplot2.tidyverse.org/reference/geom_histogram.html), but uses lines instead of bars. It’s much easier to understand overlapping lines than bars.

```{r}
freq_poly <- ggplot(diamonds, aes(x=carat, color=cut)) + 
  geom_freqpoly(show.legend = FALSE)
freq_poly
```

#### Overlapping histograms vs frequency polygons

```{r}
overlaying_histo + freq_poly
```

You can also handle overlapping histograms with the position adjustment. When bars from different groups land in the same bin, the position adjustment decides how they’re laid out. "identity" overlays them (pair it with alpha, below), "dodge" interleaves them side by side, and "stack" (the default) piles them on top of each other.

```{r}
#| out-width: "100%"

identity <- ggplot(ToothGrowth, aes(x = len, fill = supp)) +
  geom_histogram(binwidth = 3, alpha = 0.7, position = "identity", show.legend = FALSE) +
  scale_fill_viridis_d() +
  labs(x = "Tooth length", y = "Count", fill = "Supplement", title = "Identity") +
  theme_minimal()

dodge <- ggplot(ToothGrowth, aes(x = len, fill = supp)) +
  geom_histogram(binwidth = 3, position = "dodge",  show.legend = FALSE) +
  scale_fill_viridis_d() +
  labs(x = "Tooth length", y = "Count", fill = "Supplement", title = "Dodge") +
  theme_minimal()

stack <- ggplot(ToothGrowth, aes(x = len, fill = supp)) +
  geom_histogram(binwidth = 3, position = "stack", alpha = 0.7) +
  scale_fill_viridis_d() +
  labs(x = "Tooth length", y = "Count", fill = "Supplement", title = "Stack") +
  theme_minimal()

identity + dodge + stack
```

### Density plots and dotplots
Also interesting, but not for now (<https://r4ds.had.co.nz/exploratory-data-analysis.html#cat-cont>)


### Vertical histograms

Most histograms you'll see plot the continuous variable on the x-axis, but that is not a requirement.

```{r}
ggplot(diamonds, aes(y=carat)) + 
  geom_histogram(show.legend = FALSE)
```

::: {callout.warning}
This is not possible in ECharts.
:::

## Important considerations

### Zero-baseline

Histograms must be plotted with a zero-valued baseline.[^baseline]

### Absolute vs relative frequency
Depending on your visualization goals, you may want to change the units on the vertical axis of the plot as being in terms of absolute frequency or relative frequency.

- **Absolute frequency** is just the natural count of occurrences in each bin
- **Relative frequency** is the proportion of occurrences in each bin.

The choice of axis units will depend on what kinds of comparisons you want to emphasize about the data distribution.

### Where bins start is important
Two histograms can have the same bin-width and still look different. The exact look depends also on where the first bin starts.You can control this with the `boundary` parameter.

::: {.callout-note}
`boundary = 0`: "Require that one bin edge falls exactly on the value 0." The complete set of possible edges is an infinite grid:

```
..., boundary − 2·binwidth, boundary − binwidth, boundary, boundary + binwidth, boundary + 2·binwidth, ...
```

In other words, every edge is `boundary + n × binwidth` for some integer n. With `boundary = 0` and `binwidth = 1000`, that grid is:

```
..., -1000, 0, 1000, 2000, 3000, 4000, ...
```

ggplot then just keeps whichever bins in that grid are needed to cover your data.

With `boundary = 250` and `binwidth = 1000`, the grid is:

```
..., 250, 1250, 2250, 3250, 4250, ...
```
:::

```{r}
default_boundary <- ggplot(economics, aes(unemploy)) + 
  geom_histogram(binwidth=1000)

custom_boundary_1 <- ggplot(economics, aes(unemploy)) + 
  geom_histogram(binwidth=1000, boundary = 0)

custom_boundary_2 <- ggplot(economics, aes(unemploy)) + 
  geom_histogram(binwidth=1000, boundary = 250)

default_boundary + custom_boundary_1 + custom_boundary_2
```

## Questions

- handling of missing values?

- When a value is on a bin boundary, it will consistently be assigned to the bin on its right or its left (or into the end bins if it is on the end points). Which side is chosen depends on the visualization tool; some tools have the option to override their default preference.

  - how is ECharts doing this? not extremely important but more as a documentation for us

## Variable type constraints

- the variable to be visualized must be continuous: **numeric or datetime**


[^baseline]: <https://www.atlassian.com/data/charts/histogram-complete-guide>

[^bins]: <https://wzbillings.com/posts/2022-12-17_histogram-bins/>



