---
title: "Line Chart"
format:
  html:
    number-sections: true
editor: visual
---

```{r}
#| echo: false
#| warning: false

library(dplyr)
library(ggplot2)
library(patchwork)
```

**Preliminary notes**

- by default, R evaluates dates (`date` type) as continuous or interval data.

# Datasets

## Economics dataset with an additional category (`character`) column

```{r}
economics <- economics |>
 
  mutate(category = rep(c("A", "B", "C"), length.out = n()))
head(economics)
```

## Economics dataset with a category (`character`) column and multiple unemploy values for each unique row

```{r}
economics_multi <- economics |>
  slice(rep(1:n(), each = 3)) |>
  group_by(date) |>
  mutate(unemploy = unemploy + c(-50, 0, 50)) |>
  ungroup()
head(economics_multi)
```

## Simulated data: 20 random measurements per `date` over 10 days

```{r}
# Simulate raw data: ~20 measurements per day over 10 days
raw_data <- data.frame(
  date = rep(seq(as.Date("2024-01-01"), by = "day", length.out = 10), each = 20),
  response_time = rnorm(200, mean = rep(c(120, 125, 118, 130, 135, 128, 122, 119, 124, 121), each = 20), sd = 8)
)
head(raw_data)
```

# geom_line

`geom_line()` connects observations **in order of the variable on the x axis**, sorting x-values before drawing *by default*. Every line geom needs to know where one line ends and the next begins. It has to decide which points belong to the same line and get connected, and which belong to different lines and should be kept apart.

The mechanism used is the **`group`** aesthetic:

- same group value = same line

- different group value = different line

Without an explicit `group` mapping, ggplot implicitly takes *every discrete (categorical) **aesthetic** currently mapped* and **combines** them into one composite label. More precisely, it's doing something like `interaction()` of all those discrete variables, gluing their values together row by row to produce a grouping key. Put differently: ggplot's **default** `group` aesthetic is the **interaction of all discrete variables** in the plot (i.e. every unique combination of their values becomes one group)**.** All rows sharing the same **`group`** values get connected into a single line. Continuous variables are **ignored** when forming groups.

**Example:** If you have `color = region` (3 levels) and `linetype = product` (2 levels), the default grouping is the 6 combinations `region × product`. Mapping something to `x`, `y`, or a continuous `color` gradient does *not* create groups.

```{r}
ggplot(economics, aes(date, unemploy)) +
  geom_line()
```

::: {.callout-important collapse="true"}
You can set the `group` explicitly:

```{r}
ggplot(economics, aes(x = date, y = unemploy, group = category)) +
  geom_line()          # one line per region, all the same color
```

`group` tells ggplot2 how to split the data ***without*** also changing color, line type, etc.

```{r}
ggplot(economics, aes(x = date, y = unemploy, group = category)) +
  geom_line() # one line per region, all the same color
```

\
If you want both to matter, you'd use `group = interaction(region, year_factor)`.
:::

# Variable type constraints

## ***categorical vs continuous***

::: callout-note
- By default, ggplot does not connect dots from *different* categories, but only dots *within* one category.

- `chr`, `factor` share behaviour and can be seen as "categorical"

- `date` are handled specifically
:::

### unique observations per category

If each category contains only a unique observation, a warning is shown. Intuitively this makes sense: how can you meaningfully connect the dots? You're essentially saying "*connect each dot **within** one factor level*", but there is only one per factor level. A dot cannot be connected to itself.

```{r}
ggplot(economics, aes(as.factor(date), unemploy)) +
  geom_line()
```

### multiple observations

`geom_line` connects all dots **within** a category, never across different categories.

```{r}
ggplot(economics_multi, aes(as.factor(date), unemploy)) +
  geom_line()
```

Different examples for multiple values per category, but this time the x-axis is of type `chr`. You can see the same pattern: values within a category are connected, but values from different categories are not connected.

```{r}
ggplot(economics, aes(category, unemploy)) +
  geom_line()
```

::: callout-note
- **only** for categorical (`chr`, `factor`) x-axis, dates are treated differently

- different from ECharts where you can create series as you like, allowing you to connect dots from separate categories.
:::

## **continuous vs continuous**

```{r}
ggplot(economics, aes(unemploy, pop)) +
  geom_line()
```

# Additional mappings (`colour` examples)

## `chr` and `fct` colour aesthetic

`chr` variables are internally coerced into a `factor`, so the behavior is identical. In my opinion, in Adept, we should stick to `CATEGORY` and not do any internal coercion to other types.

```{r}
#| out-width: "100%"
chr_colour_mapping <- ggplot(economics, aes(date, unemploy, colour=as.character(category))) +
  geom_line()

factor_colour_mapping <- ggplot(economics, aes(date, unemploy, colour=as.factor(category))) +
  geom_line()

chr_colour_mapping + factor_colour_mapping
```

## continuous colour aesthetic

Any numeric mapped to an aesthetic gets a continuous colour scale.

::: callout-danger
I think this plot below where a mapped variable needs a continuous color scale is a real limitation in ECharts. VisualMap initally felt like the right tool, but that does not work as expected.
:::

```{r}
ggplot(economics, aes(x = date, y = unemploy, colour = pop)) +
  geom_line()
```

::: {callout-important}
### Conclusions

- ggplot2's default group is the interaction of the *discrete* aesthetics only. Continuous aesthetics never feed into grouping.

- `factor` and `chr` values are handled as discrete variables in ggplot. By default, a discrete variable mapped to a grouping aesthetic connects within its levels, not across.

- `date`s are not handled like categorical values, their handling resembles more like that of continuous variables. Dates are numeric under the hood. This is why mapping a date variable to e.g the colour aesthetic produces a continuous color scale.
:::

# Most likely use case: time series

A very common (and probably our most frequent one) use case for line charts are **time series**: date(time) *vs* some continuous quantity. Additionally,

- a **third (categorical) mapping** to create one line par unique value mapping variable value

```{r}
ggplot(economics, aes(date, unemploy, colour=category)) +
  geom_line(key_glyph="timeseries")
```

- a **summary function** to aggregate multiple y-values for a single x-value

  - Example: multiple values for a specific time (compare both plots below)

```{r}
#| echo: false

# plot 1
plot_1 <- ggplot(raw_data, aes(date, response_time)) +
  geom_line()

# plot 2
set.seed(42)

summary_data <- raw_data |>
  group_by(date) |>
  summarise(
    mean_rt = mean(response_time),
    sd_rt   = sd(response_time),
    se_rt   = sd_rt / sqrt(n()),
    .groups = "drop"
  )

summary_data

plot_2 <- ggplot(summary_data, aes(x = date, y = mean_rt)) +
  geom_ribbon(aes(ymin = mean_rt - sd_rt, ymax = mean_rt + sd_rt), 
              fill = "steelblue", alpha = 0.2) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  labs(x = "Date", y = "Response Time (ms)",
       title = "Daily Average Response Time ± 1 SD") +
  theme_minimal()

# works implicitely too
# ggplot(raw_data, aes(x = date, y = response_time)) +
#   stat_summary(fun = mean, geom = "line", color = "steelblue", linewidth = 1) +
#   stat_summary(fun = mean, geom = "point", color = "steelblue", size = 2) +
#   stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1), 
#                geom = "ribbon", alpha = 0.2, fill = "steelblue") +
#   labs(x = "Date", y = "Response Time (ms)",
#        title = "Daily Average Response Time ± 1 SD (computed by ggplot)") +
#   theme_minimal()

plot_1 + plot_2
```

## Important considerations

- x and y axis can theoretically be flipped \[not sure if we should provide this ability\]

- the use case for categorical (non-date) variables on the x-axis is so limited (I cannot think of one), that I think it's worth considering to **not** support them in Adept. This way we can focus on time vs continuous (+ additional dimension mappings).

- chart display options should include:

  - for the line itself: alpha, linetype, linewidth
  - for the legend: timeseries keyglyph, color palette (probably inherited from a global palette)

- **TBD:** `N/A` handling (check echarts default behaviour; in my opinion a gap is the best choice). Otherwise maybe a toggle to include/exclude `NA`s?

- don't enforce a zero baseline (there is a control for this in ECharts)

- no curved/smoothed interpolation between points: line charts connect dots in **straight lines**.

- summary functions: useful when there is no 1:1 mapping between x and y (multiple y's for a single x)

  - can we pre-compute a `group_by` x-variable to know how many y-values there are per x-variable group? In case there are non-unique ones ones (multiple observations for a signel date), we could add a message or hint.

  - option to show a shaded confidence interval

  - check how to do this in ECharts: for mean+uncertainty we probably need a function that

    1.  groups variables by unique x-axis values
    2.  computes mean per x-axis "category"
    3.  computes confidence interval per x-axis "category"
    4.  outputs per category: mean value, upper level, lower level

  - like `stat_summary(fun.data = mean_sdl)` in ggplot

  - check how much of this is already implemented in the `/stats` endpoint

## Zoom on the summary function feature

### Multiple y-values per date {#multi-y-per-date}

In the plot below, we plot date vs response_time (continuous). For each unique x value, there are multiple associated y values.

```{r}
ggplot(raw_data, aes(date, response_time)) +
  geom_line()
```

```{r}

```

Introducing a summary function to summarize individual points per unique x-axis is the better choice here. In ggplot you can do this without any explicit data transformations using `stat_summary`:

::: callout-note
`stat_summary` allows for tremendous flexibility in the specification of summary functions. The summary function can either operate on a data frame (with argument name **`fun.data`**) or on a vector (**`fun.y`**, **`fun.ymax`**, **`fun.ymin`**).

The default behavior is to **split the data into groups by every distinct value of `x`**, then apply your summary function (`fun` or `fun.data`) separately to the `y`-values within each group.
:::

```{r}
ggplot(raw_data, aes(x = date, y = response_time)) +
  stat_summary(fun = mean, geom = "line", color = "steelblue", linewidth = 1) +
  #stat_summary(fun = mean, geom = "point", color = "steelblue", size = 2) +
  stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1), 
               geom = "ribbon", alpha = 0.2, fill = "steelblue") +
  labs(x = "Date", y = "Response Time (ms)",
       title = "Daily Average Response Time ± 1 SD (computed by ggplot)") +
  theme_minimal()
```

# Edge cases

- Mapping (not x or y) to continuous variables. Hard to reproduce in ECharts (and probably not scalable with increasing data points). You need a custom renderer to color individual pixels.

```{r}
ggplot(economics, aes(date, unemploy, colour=pop)) +
  geom_line(key_glyph="timeseries")
```

- Mapping (not x or y) to date variables. ggplot uses a specialized continuous scale. Remember that dates are numeric under the hood.

```{r}
ggplot(economics, aes(date, unemploy, colour=date)) +
  geom_line(key_glyph="timeseries")
```

- Mapping (not x or y) to categorical variables with multiple values per category

<!-- -->

- Multiple mappings (not x or y) (4th, 5th, dimension...)

- Mapping (not x or y) to categorical variables with too many groups. What should be the cutoff. Note that below you only see the legend.

```{r}
ggplot(economics, aes(date, unemploy, colour=as.factor(date))) +
  geom_line(key_glyph="timeseries")
```

- `N/A` handling

## Questions to answer

- How to format the time axis labels? our date objects are all iso8601 on the client side even if they only contains a time component... in that case showing only the time is better? or let echarts run its default parser?

  - check the raw_data dataset in Adept and plot date vs response time

# Compatibility Table

## X and Y

+------------------------------------------+-------------+-------------------------+-------------------------+-----------+
| X                                        | Y           | Supported (technically) | Supported (practically) | Notes     |
+==========================================+=============+=========================+=========================+===========+
| categorical                              | categorical | NO                      | NO                      |           |
+------------------------------------------+-------------+-------------------------+-------------------------+-----------+
| categorical (1 obs. per category)        | continuous  | YES                     | NO                      |           |
+------------------------------------------+-------------+-------------------------+-------------------------+-----------+
| categorical (multiple obs. per category) | continuous  | YES                     | NO                      |           |
+------------------------------------------+-------------+-------------------------+-------------------------+-----------+
| continuous                               | continuous  | **YES**                 | **YES**                 |           |
+------------------------------------------+-------------+-------------------------+-------------------------+-----------+
| time                                     | continuous  | **YES**                 | **YES**                 |           |
+------------------------------------------+-------------+-------------------------+-------------------------+-----------+

## Additional mappings

+-------------+-------------------------+-------------------------+----------------------------------------------------------------------+
| Mapping     | Supported (technically) | Supported (practically) | Note                                                                 |
+=============+=========================+=========================+======================================================================+
| categorical | **YES**                 | **YES**                 |                                                                      |
+-------------+-------------------------+-------------------------+----------------------------------------------------------------------+
| datetime    | YES                     | YES                     | ggplot: special continuous scale                                     |
|             |                         |                         |                                                                      |
|             |                         |                         | adept (currently): categorical scale, one series per unique datetime |
+-------------+-------------------------+-------------------------+----------------------------------------------------------------------+
| continuous  | YES                     | **NO**                  | ECharts limitation (?)                                               |
+-------------+-------------------------+-------------------------+----------------------------------------------------------------------+
