---
title: "Line Chart"
format: html
editor: visual
---

```{r}
library(dplyr)
library(ggplot2)
library(patchwork)
```

**Note: By default, R evaluates dates as continuous or interval data.**

```{r}
economics <- economics |>
  mutate(date_factor = as.factor(date)) |> 
  mutate(category = rep(c("A", "B", "C"), length.out = n()))
head(economics)
```

# geom_line

`geom_line()` connects observations **in order of the variable on the x axis**, sorting x-values before drawing by default.

Every line geom needs to know where one line ends and the next begins, so you cannot just provide a bunch of points to `geom_line()`. It has to decide which points belong to the same line and get connected, and which belong to different lines and should be kept apart. The mechanism used is the **`group`** aesthetic:

- same group value = same line

- different group value = different line

`group` answers the question "which line does this row belong to?"

When you don't map `group` explicitly ggplot does it implicitly and the rule it uses is: *take every discrete (categorical) **aesthetic** currently mapped, and combine them into one composite label.* More precisely, it's doing something like `interaction()` of all those discrete variables, gluing their values together row by row to produce a grouping key.

Phrased differently: **ggplot's default `group` aesthetic is the interaction of all discrete variables in the plot.**

```{r}
ggplot(economics, aes(date, unemploy)) +
  geom_line()
```

### Variable type constraints

Do both axis have some constraint regarding the type (categorical, continuous)?

- ***categorical (unique observations) vs continuous***

  If there is only one observation per "category" you get a warning. This makes sense, because how is ggplot supposed to connect these dots for it to make sense? You're essentially saying "*connect each dot within one factor level*", but there is only one per factor level and a dot cannot be connected to itself.

```{r}
ggplot(economics, aes(date_factor, unemploy)) +
  geom_line()
```

- ***categorical (multiple observations) vs continuous***

  By default, geom_line now connects dots with lines for all observations within **each** factor level. Observations from different factor levels are never connected.

```{r}
economics_multi <- economics |>
  slice(rep(1:n(), each = 3)) |>
  group_by(date) |>
  mutate(unemploy = unemploy + c(-50, 0, 50)) |>
  ungroup()
head(economics_multi)
```

```{r}
ggplot(economics_multi, aes(date_factor, unemploy)) +
  geom_line()
```

- **continuous vs continuous**

```{r}
ggplot(economics, aes(unemploy, pop, color=date)) +
  geom_line()
```

### 

### Most likely use case

The classical use case will be **time series**: time *vs* some continuous quantity. Additionally,

- a third mapping to split the line chart into line charts per unique value in the mapping. Not sure how to call this mapping (colour?, group_by?)

```{r}
ggplot(economics, aes(date, unemploy, colour=category)) +
  geom_line(key_glyph="timeseries")
```

- a summary function to apply in case there are multiple y-values for a single x-value

  - e.g. multiple values for a specific time (compare both plots below)

```{r}
# Simulate raw data: ~20 measurements per day over 10 days
raw_data <- data.frame(
  date = rep(seq(as.Date("2024-01-01"), by = "day", length.out = 10), each = 20),
  response_time = rnorm(200, mean = rep(c(120, 125, 118, 130, 135, 128, 122, 119, 124, 121), each = 20), sd = 8)
)
```

```{r}
# plot 1
plot_1 <- ggplot(raw_data, aes(date, response_time)) +
  geom_line()

# plot 2
set.seed(42)

summary_data <- raw_data |>
  group_by(date) |>
  summarise(
    mean_rt = mean(response_time),
    sd_rt   = sd(response_time),
    se_rt   = sd_rt / sqrt(n()),
    .groups = "drop"
  )

summary_data

plot_2 <- ggplot(summary_data, aes(x = date, y = mean_rt)) +
  geom_ribbon(aes(ymin = mean_rt - sd_rt, ymax = mean_rt + sd_rt), 
              fill = "steelblue", alpha = 0.2) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  labs(x = "Date", y = "Response Time (ms)",
       title = "Daily Average Response Time ± 1 SD") +
  theme_minimal()

# works implicitely too
# ggplot(raw_data, aes(x = date, y = response_time)) +
#   stat_summary(fun = mean, geom = "line", color = "steelblue", linewidth = 1) +
#   stat_summary(fun = mean, geom = "point", color = "steelblue", size = 2) +
#   stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1), 
#                geom = "ribbon", alpha = 0.2, fill = "steelblue") +
#   labs(x = "Date", y = "Response Time (ms)",
#        title = "Daily Average Response Time ± 1 SD (computed by ggplot)") +
#   theme_minimal()

plot_1 + plot_2
```

## Important considerations

- x and y axis can theoretically be flipped (if we want to support this is another question)

- for line charts in Adept, categorical values should not be supported

  - only time and continuous

- chart display options to include:

  - for the line itself: alpha, linetype, linewidth

  - for the legend: timeseries keyglyph, color palette (probably use the global one)

- **TBD:** `N/A` handling (check echarts default behaviour; in my opinion a gap is the best choice)

  - otherwise maybe a toggle to include/exclude NAs?

- no enforced zero baseline if we don't have to

  - check how echarts calculates the scale

- no curved/smoothed interpolation between points

  - line charts should connect dots in **straight lines**

- summary functions to use if there is not a 1:1 mapping between x-and y

  - maybe even precompute the number of observations per x-axis factor: if there are non-unique ones, add a message/hint to add

  - mean, median, ...

  - option to show a shaded confidence interval

  - check how to do this in echarts: for mean+uncertainty we probably need a function that

    1.  groups variables by unique x-axis values
    2.  computes mean per x-axis "category"
    3.  computes confidence interval per x-axis "category"
    4.  outputs per category: mean value, upper level, lower level

  - like `stat_summary(fun.data = mean_sdl)` in ggplot

## Zoom on the summary function feature

In the plot below, we plot time vs continuous (x vs y) data where for each unique x value, there are multiple associated y values.

```{r}
ggplot(raw_data, aes(date, response_time)) +
  geom_line()
```

Introducing a summary function to somehow summarize the indiviual points per unique x-axis is probably much more suited here. You can do this directly in ggplot without the explicit data transformations:

**`stat_summary`** allows for tremendous flexibility in the specification of summary functions. The summary function can either operate on a data frame (with argument name **`fun.data`**) or on a vector (**`fun.y`**, **`fun.ymax`**, **`fun.ymin`**).

- Internally, its default behavior is: **split the data into groups by every distinct value of `x`**, then apply your summary function (`fun` or `fun.data`) separately to the `y`-values within each group.

```{r}
ggplot(raw_data, aes(x = date, y = response_time)) +
  stat_summary(fun = mean, geom = "line", color = "steelblue", linewidth = 1) +
  stat_summary(fun = mean, geom = "point", color = "steelblue", size = 2) +
  stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1), 
               geom = "ribbon", alpha = 0.2, fill = "steelblue") +
  labs(x = "Date", y = "Response Time (ms)",
       title = "Daily Average Response Time ± 1 SD (computed by ggplot)") +
  theme_minimal()
```

# Compatibility Table

| X | Y | Supported (technically) | Supported (practically) | Notes |
|---------------------------|-----------|-------------|------------|------------------|
| categorical | categorical | NO | NO |  |
| categorical (1 obs. per category) | continuous | YES | NO |  |
| categorical (multiple obs. per category) | continuous | YES | NO |  |
| continuous | continuous | **YES** | **YES** |  |
| time | continuous | **YES** | **YES** |  |
