---
title: "Line Chart"
format:
  html:
    number-sections: true
editor: visual
---

```{r}
#| echo: false
#| warning: false

library(dplyr)
library(ggplot2)
library(patchwork)
```

**Preliminary notes**

- by default, R evaluates dates (`date` type) as continuous or interval data.

# Datasets

## [Economics dataset with an additional category (`character`) column and a `date_factor`]{.underline} [column]{.underline}

```{r}
economics <- economics |>
  mutate(date_factor = as.factor(date)) |> 
  mutate(category = rep(c("A", "B", "C"), length.out = n()))
head(economics)
```

## Economics dataset [with a category (`character`) column and multiple unemploy values for each unique row]{.underline}

```{r}
economics_multi <- economics |>
  slice(rep(1:n(), each = 3)) |>
  group_by(date) |>
  mutate(unemploy = unemploy + c(-50, 0, 50)) |>
  ungroup()
head(economics_multi)
```

## [Simulated raw data: 20 measurements per `date` over 10 days]{.underline}

```{r}
# Simulate raw data: ~20 measurements per day over 10 days
raw_data <- data.frame(
  date = rep(seq(as.Date("2024-01-01"), by = "day", length.out = 10), each = 20),
  response_time = rnorm(200, mean = rep(c(120, 125, 118, 130, 135, 128, 122, 119, 124, 121), each = 20), sd = 8)
)
head(raw_data)
```

# geom_line

`geom_line()` connects observations **in order of the variable on the x axis**, sorting x-values before drawing *by default*. Every line geom needs to know where one line ends and the next begins, so you cannot just provide a bunch of points to `geom_line()`. It has to decide which points belong to the same line and get connected, and which belong to different lines and should be kept apart.

The mechanism used is the **`group`** aesthetic:

- same group value = same line

- different group value = different line

Without an explicit `group` mapping, ggplot implicitly takes *every discrete (categorical) **aesthetic** currently mapped* and **combines** them into one composite label. More precisely, it's doing something like `interaction()` of all those discrete variables, gluing their values together row by row to produce a grouping key. Phrased differently: **ggplot's default `group` aesthetic is the interaction of all discrete variables in the plot (**i.e. every unique combination of their values becomes one group)**.** All rows sharing the same **`group`** values get connected into a single line.

Example: If you have `color = region` (3 levels) and `linetype = product` (2 levels), the default grouping is the 6 combinations `region × product`. C**ontinuous variables are ignored** when forming groups. Mapping something to `x`, `y`, or a continuous `color` gradient does *not* create groups.

```{r}
ggplot(economics, aes(date, unemploy)) +
  geom_line()
```

::: {.callout-important collapse="true"}
You can set `group` explicitly:

```{r}
ggplot(economics, aes(x = date, y = unemploy, group = category)) +
  geom_line()          # one line per region, all the same color
```

`group` tells ggplot2 how to split the data ***without*** also changing color, linetype, etc. If you set `group = region` explicitly, that **overrides** the interaction default entirely — ggplot2 uses your `group` and stops auto-combining the discrete aesthetics. That's occasionally surprising: mapping `color` to something discrete no longer subdivides the lines once you've pinned `group`. If you want both to matter, you'd use `group = interaction(region, year_factor)`.
:::

# Variable type constraints

## ***categorical vs continuous***

### unique observations

For unique observations per "category", a warning is shown. Intuitively this makes sense: how is ggplot supposed to meaningfully connect these dots? You're essentially saying "*connect each dot within one factor level*", but there is only one per factor level. A dot cannot be connected to itself.

```{r}
ggplot(economics, aes(date_factor, unemploy)) +
  geom_line()
```

### multiple observations

By default, `geom_line` now connects dots for all observations **within** each factor level. Observations from different factor levels are never connected.

```{r}
ggplot(economics_multi, aes(date_factor, unemploy)) +
  geom_line()
```

## **continuous vs continuous**

Same default behavior: the x-axis is sorted.

```{r}
ggplot(economics, aes(unemploy, pop)) +
  geom_line()
```

# Additional mappings (`colour` examples)

## `chr` and `fct` colour aesthetic

`chr` variables are internally coerced into a `factor`, so the behavior is identical. In Adept, we should stick to `CATEGORY` in my opinion and not do any internal coercion to other types.

```{r}
#| out-width: "100%"
chr_colour_mapping <- ggplot(economics, aes(date, unemploy, colour=as.character(category))) +
  geom_line()

factor_colour_mapping <- ggplot(economics, aes(date, unemploy, colour=as.factor(category))) +
  geom_line()

chr_colour_mapping + factor_colour_mapping
```

## continuous colour aesthetic

Any numeric mapped to an aesthetic gets a continuous colour scale.

::: callout-danger
I think this plot below where a mapped variable needs a continuous color scale is a real limitation in ECharts. VisualMap initally felt like the right tool, but that does not work as expected.
:::

```{r}
ggplot(economics, aes(x = date, y = unemploy, colour = pop)) +
  geom_line()
```

# Most likely use case: time series

A very common (and probably the most frequent) use case for line charts are time series: date(time) *vs* some continuous quantity. Additionally,

- a **third (categorical) mapping** to create one line par unique value mapping variable value

```{r}
ggplot(economics, aes(date, unemploy, colour=category)) +
  geom_line(key_glyph="timeseries")
```

- a **summary function** to aggregate multiple y-values for a single x-value

  - Example: multiple values for a specific time (compare both plots below)

```{r}
#| echo: false

# plot 1
plot_1 <- ggplot(raw_data, aes(date, response_time)) +
  geom_line()

# plot 2
set.seed(42)

summary_data <- raw_data |>
  group_by(date) |>
  summarise(
    mean_rt = mean(response_time),
    sd_rt   = sd(response_time),
    se_rt   = sd_rt / sqrt(n()),
    .groups = "drop"
  )

summary_data

plot_2 <- ggplot(summary_data, aes(x = date, y = mean_rt)) +
  geom_ribbon(aes(ymin = mean_rt - sd_rt, ymax = mean_rt + sd_rt), 
              fill = "steelblue", alpha = 0.2) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  labs(x = "Date", y = "Response Time (ms)",
       title = "Daily Average Response Time ± 1 SD") +
  theme_minimal()

# works implicitely too
# ggplot(raw_data, aes(x = date, y = response_time)) +
#   stat_summary(fun = mean, geom = "line", color = "steelblue", linewidth = 1) +
#   stat_summary(fun = mean, geom = "point", color = "steelblue", size = 2) +
#   stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1), 
#                geom = "ribbon", alpha = 0.2, fill = "steelblue") +
#   labs(x = "Date", y = "Response Time (ms)",
#        title = "Daily Average Response Time ± 1 SD (computed by ggplot)") +
#   theme_minimal()

plot_1 + plot_2
```

## Important considerations

- x and y axis can theoretically be flipped \[not sure if we should provide this ability\]

- the use case for categorical (non-date) variables on the x-axis is so limited (I cannot think of one), that I think it's worth considering to **not** support them in Adept. This way we can focus on time vs continuous (+ additional dimension mappings).

- chart display options should include:

  - for the line itself: alpha, linetype, linewidth
  - for the legend: timeseries keyglyph, color palette (probably inherited from a global palette)

- **TBD:** `N/A` handling (check echarts default behaviour; in my opinion a gap is the best choice). Otherwise maybe a toggle to include/exclude `NA`s?

- don't enforce a zero baseline (there is a control for this in ECharts)

- no curved/smoothed interpolation between points: line charts connect dots in **straight lines**.

- summary functions: useful when there is no 1:1 mapping between x and y (multiple y's for a single x)

  - can we pre-compute a `group_by` x-variable to know how many y-values there are per x-variable group? In case there are non-unique ones ones (multiple observations for a signel date), we could add a message or hint.

  - option to show a shaded confidence interval

  - check how to do this in ECharts: for mean+uncertainty we probably need a function that

    1.  groups variables by unique x-axis values
    2.  computes mean per x-axis "category"
    3.  computes confidence interval per x-axis "category"
    4.  outputs per category: mean value, upper level, lower level

  - like `stat_summary(fun.data = mean_sdl)` in ggplot

  - check how much of this is already implemented in the `/stats` endpoint

## Zoom on the summary function feature

### Multiple y-values per date {#multi-y-per-date}

In the plot below, we plot date vs response_time (continuous). For each unique x value, there are multiple associated y values.

```{r}
ggplot(raw_data, aes(date, response_time)) +
  geom_line()
```

Introducing a summary function to summarize individual points per unique x-axis is the better choice here. In ggplot you can do this without any explicit data transformations using `stat_summary`:

::: callout-note
`stat_summary` allows for tremendous flexibility in the specification of summary functions. The summary function can either operate on a data frame (with argument name **`fun.data`**) or on a vector (**`fun.y`**, **`fun.ymax`**, **`fun.ymin`**).

The default behavior is to **split the data into groups by every distinct value of `x`**, then apply your summary function (`fun` or `fun.data`) separately to the `y`-values within each group.
:::

```{r}
ggplot(raw_data, aes(x = date, y = response_time)) +
  stat_summary(fun = mean, geom = "line", color = "steelblue", linewidth = 1) +
  #stat_summary(fun = mean, geom = "point", color = "steelblue", size = 2) +
  stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1), 
               geom = "ribbon", alpha = 0.2, fill = "steelblue") +
  labs(x = "Date", y = "Response Time (ms)",
       title = "Daily Average Response Time ± 1 SD (computed by ggplot)") +
  theme_minimal()
```

# Edge cases

- Mapping (not x or y) to continuous variables. Hard to reproduce in ECharts (and probably not scalable with increasing data points). You need a custom renderer to color individual pixels.

```{r}
ggplot(economics, aes(date, unemploy, colour=pop)) +
  geom_line(key_glyph="timeseries")
```

- Mapping (not x or y) to date variables. ggplot uses a specialized continuous scale.

```{r}
ggplot(economics, aes(date, unemploy, colour=date)) +
  geom_line(key_glyph="timeseries")
```

- Mapping (not x or y) to categorical variables with multiple values per category

```{r}
ggplot(raw_data, aes(date, response_time)) +
  geom_line()
```

- multiple mappings (not x or y) (4th, 5th, dimension...)

- Mapping (not x or y) to categorical variables with too many groups. What should be the cutoff. Note that below you only see the legend.

```{r}
ggplot(economics, aes(date, unemploy, colour=date_factor)) +
  geom_line(key_glyph="timeseries")
```

- how to handle `N/A`s

# Compatibility Table

## X and Y

| X | Y | Supported (technically) | Supported (practically) | Notes |
|---------------|---------------|---------------|---------------|---------------|
| categorical | categorical | NO | NO |  |
| categorical (1 obs. per category) | continuous | YES | NO |  |
| categorical (multiple obs. per category) | continuous | YES | NO |  |
| continuous | continuous | **YES** | **YES** |  |
| time | continuous | **YES** | **YES** |  |

## Additional mappings

| Mapping | Supported (technically) | Supported (practically) | Note |
|----|----|----|----|
| categorical | **YES** | **YES** |  |
| continuous | YES | **NO** | ECharts limitation (?) |
