---
title: "Line Chart"
format: html
editor: visual
---

**Note: By default, R evaluates dates as continuous or interval data.**

```{r}
economics <- economics |>
  mutate(date_factor = as.factor(date)) |> 
  mutate(category = rep(c("A", "B", "C"), length.out = n()))
head(economics)
```

# geom_line

`geom_line()` connects observations **in order of the variable on the x axis**. `geom_line()` sorts by the x value before drawing (!!!).

**Every line geom needs to know where one line ends and the next begins.** You cannot just hand `geom_line()` a pile of points. It has to decide which points belong to the same line and get connected, and which belong to different lines and should be kept apart.

The mechanism it uses for that decision is the **`group`** aesthetic: same group value = same line; different group value = different line. `group` answers the question "which line does this row belong to?".

When you don't map `group` explicitly, ggplot has to invent one. The rule it uses is: *take every discrete (categorical) **aesthetic** currently mapped, and combine them into one composite label.* Concretely, it's doing something like `interaction()` of all those discrete variables, gluing their values together row by row to produce a grouping key.

```{r}
ggplot(economics, aes(date, unemploy)) +
  geom_line()
```

The `group` aesthetic determines which cases are connected together.

### Variable type constraints

Do both axis have some constraint regarding the type (categorical, continuous)?

- *categorical (unique observations) vs continuous*

  If only one observation per "category" you get a warning. This makes sense: how would you connect these dots so that it makes logically sense? ggplot basically says: "i am supposed to connect each dot within one factor level, but there is only one per factor level? I can not connect a dot to itself.

**ggplot's default `group` aesthetic is the interaction of all discrete variables in the plot.**

```{r}
ggplot(economics, aes(date_factor, unemploy)) +
  geom_line()
```

- *categorical (multiple observations) vs continuous*

```{r}
economics_multi <- economics |>
  slice(rep(1:n(), each = 3)) |>
  group_by(date) |>
  mutate(unemploy = unemploy + c(-50, 0, 50)) |>
  ungroup()
```

**ggplot's default `group` aesthetic is the interaction of all discrete variables in the plot.** `geom_line()` now draws a separate line **within each factor level**, and never connects across levels:

```{r}
ggplot(economics_multi, aes(date_factor, unemploy)) +
  geom_line()
```

- continuous vs continuous

```{r}
ggplot(economics, aes(unemploy, unemploy, color=date)) +
  geom_line()
```

### 

### Most likely use case

The classical use case will be **time series**: time (as a continuous variable) on the x-axis and some continuous quantity. A third dimension to split the line charts will also be very useful.

```{r}
ggplot(economics, aes(date, unemploy, colour=category)) +
  geom_line(key_glyph="timeseries")
```

- Note: x and y can be flipped

- time is not a separate "case": decide if time is handled as categorical or continuous and the table below holds

- for supported combinations, user should be able to "group" by a third (categorical!!!) variable: not sure how we should call this. colour could work, but group_by and then a palette may be better?

- chart display options should include: alpha, linetype, linewidth

- if echarts allows this: legend should have this timeseries keyglyph

- TBD: N/A handling! check echarts, a gap in the line is not a problem to me?

| X | Y | Supported | Note |
|------------------|------------------|------------------|------------------|
| categorical | categorical | NO |  |
| categorical (1 observation per category) | continuous | YES |  |
| categorical (multiple observations per category) | continuous | YES |  |
| continuous (time!) | continuous | **YES** (most common use case) |  |
